---
title: "Peer-graded Assignment: Course Project 1"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


###1.  Code for reading in the dataset and/or processing the data

```{r echo=TRUE}
#load packages
library(dplyr)
library(ggplot2)
library(gridExtra)

#import data
activityData<-read.csv("activity.csv")
#formate date column as Date class
activityData$date<-as.Date(activityData$date)

```

###2.  Histogram of the total number of steps taken each day

```{r echo=TRUE}
#filter data and sum steps by data remove na
byDateData<-activityData %>% 
     group_by(date) %>%
     summarize(steps= sum(steps)) %>%
     filter(is.na(steps)==FALSE)

#generate hisotgram
ggplot(byDateData, aes(x=date, y=steps)) +
     geom_col() +
     labs(title = "steps by day", y="sum of steps")
```

###3.  Mean and median number of steps taken each day
+ mean number of steps per day
```{r, echo=TRUE}
print(mean(byDateData$steps))
```

+ median number of steps per day
```{r, echo=TRUE}
print(median(byDateData$steps))
```

###4.  Time series plot of the average number of steps taken

```{r echo=TRUE}
#filter data remove NA, group by interval and sumarize the mean number of steps
byIntDataMean<-activityData %>% 
     filter(is.na(steps)==FALSE) %>%
     group_by(interval) %>%
     summarize(steps= mean(steps)) 

#plot steps by interval
ggplot(byIntDataMean, aes(x=interval, y=steps)) +
     geom_line() + 
     labs(title = "steps per 5 minute interval", y="mean steps")

```

###5.  The 5-minute interval that, on average, contains the maximum number of steps

```{r, echo=TRUE}
topInterval<-byIntDataMean %>%
    top_n(1, steps)

print(topInterval)
```

###6.  Code to describe and show a strategy for imputing missing data
imputed the missing data by replacing NA values with the mean of the non-NA values for that interval across all timepoints.

```{r, echo=TRUE}

impData<-activityData %>% 
     group_by(interval) %>%
     mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps)) 

```

###7.  Histogram of the total number of steps taken each day after missing values are imputed

```{r, echo=TRUE}
#filter group by date and sumarize the sum of steps

byDateImpData<-impData %>% 
     group_by(date) %>%
     summarize(steps= sum(steps))

#plot sum of steps by date
ggplot(byDateImpData, aes(x=date, y=steps)) +
     geom_col() + 
     labs(title = "Imputed steps by day", y="sum of steps")

```

###8.  Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r, echo=TRUE}
#create a column of days using mutate with weekdays function, remove NAs, filter by week days, group by interval and summarize the mean steps
byIntDataMeanWD<-activityData %>% 
     mutate(day=weekdays(date)) %>%
     filter(is.na(steps)==FALSE) %>%
     filter(day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")==TRUE) %>%
     group_by(interval) %>%
     summarize(steps= mean(steps)) 

#store plot mean of steps by interval for weekdays
WDplot<-ggplot(byIntDataMeanWD, aes(x=interval, y=steps)) +
     geom_line() + 
     labs(title = "Weekdays", y="mean steps")

#create a column of days using mutate and weekdays function, remove NA, filter by week ends, group by interval and summarize the mean steps
byIntDataMeanWE<-activityData %>% 
     mutate(day=weekdays(date)) %>%
     filter(is.na(steps)==FALSE) %>%
     filter(day %in% c("Saturday", "Sunday")==TRUE) %>%
     group_by(interval) %>%
     summarize(steps= mean(steps)) 

#store plot mean of steps by interval for weekends
WEplot<-ggplot(byIntDataMeanWE, aes(x=interval, y=steps)) +
     geom_line()+ 
     labs(title = "Weekends", y="mean steps")

#return plots in with grid arrange function from gridExtra library
grid.arrange(WDplot, WEplot, ncol=2)
```



